Lemma 1

### 2021-03-04 notes/jessie-russian-lemma.tex
Updating directional derivative lemma

% alphabetical order, by convention
\newcommand{\B}{\mathcal{B}}
\newcommand{\C}{\mathcal{C}}
\newcommand{\D}{\mathcal{D}}
\newcommand{\E}{\mathbb{E}}
\newcommand{\F}{\mathcal{F}}
\newcommand{\I}{\mathcal{I}}
\newcommand{\N}{\mathcal{N}}
\renewcommand{\P}{\mathcal{P}}
\newcommand{\R}{\mathcal{R}}
\newcommand{\Sc}{\mathcal{S}}
\newcommand{\U}{\mathcal{U}}
\newcommand{\X}{\mathcal{X}}
\newcommand{\Y}{\mathcal{Y}}

@@

We define $\E_p \partial L(u,Y) = \{x \in \reals^d : x = \E_p V(Y), V(y) \in \partial L(u,y)\, p\textrm{-almost surely}, V \text{ measurable}\}$.

\begin{lemma}
	For any $z \in \reals^d$, there exists a $v \in \E_p \partial L(u,Y)$ such that $\inprod{z}{v'} = \E_p L'(u,Y,z)$.
\end{lemma}
\begin{proof}

Fix $z$.
Consider the set-valued function $Q_z : \Omega \to 2^{\reals^d}$ defined by $Q_z(\omega) := \{v' \in \reals^d : \inprod{z}{v'} = L'(u,\omega;z) \}$.
Now, we want to take a measurable function $V$ that is a selection of $Q_z$ satisfying the following two conditions:
\begin{enumerate}
	\item $\E_p V \in \E_p \partial L(u,Y)$.
	\item $\inprod{z}{\E_p V} = \E_p L'(u,Y,z)$
\end{enumerate} 
If these are true, then we can take $v := \E_p V$ and observe the result.

Observe that $Q_z$ is solid-valued and therefore measurable iff $Q_z^{-1}(x) \in \B(\Omega)$ for all $x \in \reals^d$. \jessie{Rockafellar VA Example 14.7}.
\jessie{Assumptions that we're happy with} implies then that $Q_z$ is measurable; in turn, by \jessie{Rockafellar VA 14.6}, we have a measurable selection $V : \Omega \to \reals^d$ such that $\inprod{z}{V(\omega)} = L'(u,\omega; z)$ for all $\omega \in \Omega$.

First, we want to show $\E_p V \in \E_p \partial L(u,Y)$.
That is, $\int V(\omega) dp\omega \in \int \partial L(u,\omega) dp\omega = \{\E_p V(Y) : V \textrm{ measurable, }\, V(\omega) \in \partial L(u,\omega) \, p\textrm{-almost surely}\}$.
Since $V$ is measurable, it is just left to show $V(\omega) \in \partial L(u,\omega)$ $p$-almost surely.
By construction, for all $\omega \in \Omega$, we have $\inprod{V(\omega)}{z} = L'(u,\omega;z)$.
This implies 
\begin{align*}
\inprod{V(\omega)}{z} &= L'(u,\omega;z)\\
\inprod{V(\omega)}{z} &= \sup_{x \in \partial L(u,\omega)} \inprod{x}{z} \jessie{\textrm{Boyd and Vandenberghe slides}}\\
\implies \inprod{V(\omega)}{z} &\geq \inprod{x}{z} \; \forall x \in \partial L(u,\omega)\\
z := \omega - t \text{\footnotemark}\implies \inprod{V(\omega)}{\omega - t} &\geq \inprod{x}{\omega - t} \geq L(u,\omega) - L(u,t) \; \forall x \in \partial L(u,\omega)\\
\implies \inprod{V(\omega)}{\omega-t} &\geq L(u,\omega) - L(u,t)\,\forall t \in \reals^d\\
V(\omega) &\in \partial L(u,\omega) \\ 
\end{align*}
\footnotetext{can we do this?  since $z$ is fixed but $t$ is a variable}
\jessie{This is actually a bit stronger than we need; is that part of my issue?}

Second, we want to show $\inprod{z}{v} = \E_p L'(u,Y,z)$.
This follows from linearity of expectation and construction of $V$; to see this, consider
\begin{align*}
\inprod{z}{v} &= \inprod{z}{\E_p V} \\
 &= \E_p \inprod{z}{V} \\
 &= \int \inprod{z}{V(\omega)} dp\omega \\
 &= \int L'(u,\omega; z) dp\omega \\
 &= \E_p L'(u,Y;z)
\end{align*}

>>>>

% alphabetical order, by convention
\newcommand{\B}{\mathcal{B}}
\newcommand{\C}{\mathcal{C}}
\newcommand{\D}{\mathcal{D}}
\newcommand{\E}{\mathbb{E}}
\newcommand{\F}{\mathcal{F}}
\newcommand{\I}{\mathcal{I}}
\newcommand{\N}{\mathcal{N}}
\renewcommand{\P}{\mathcal{P}}
\newcommand{\R}{\mathcal{R}}
\newcommand{\Sc}{\mathcal{S}}
\newcommand{\U}{\mathcal{U}}
\newcommand{\V}{\mathcal{V}}
\newcommand{\X}{\mathcal{X}}
\newcommand{\Y}{\mathcal{Y}}

@@

We define $\E_p \partial L(u,Y) = \{x \in \reals^d : x = \E_p V(Y), V(y) \in \partial L(u,y), V \text{ measurable}\}$.

Moreover, let $\V_{u} = \{V : \Y \to \reals^d : V \text{ measurable}, V(\omega) \in \partial L(u,\omega)\}$.

\begin{lemma}
	For any $z \in \reals^d$, there exists a $v \in \E_p \partial L(u,Y)$ such that $\inprod{z}{v} = \E_p L'(u,Y;z)$.
\end{lemma}
\begin{proof}
	\jessie{Meeting notes: - Life will be easier if you forgot/redefined expected value of the subgradient -Want $Q_z$ to be subgradients, and get the everywhere selection - Careful about pulling out expectation in bullet 2; when is inner product of RV = expectation of inner product? - Point out contradicting the second line would yield contradiction of subgradient inequality - Argue $Q_z$ is measurable - Go slowly through the definition of measurable}

Fix $z$.
Consider the set-valued function $Q_z : \Omega \to 2^{\reals^d}$ defined by $Q_z(\omega) := \{V \in \V_u: \inprod{z}{V(\omega)} = L'(u,\omega;z) \}$.
Now, we want to take a measurable function $V \in \V_u$ that is a selection of $Q_z$ satisfying the following conditions:
\begin{enumerate}
	\item $V \in \V_u$.
	\item $\E_p V \in \E_p \partial L(u,Y)$.
	\item $\inprod{z}{\E_p V} = \E_p L'(u,Y,z)$
\end{enumerate} 
If these are true, then we can take $v := \E_p V$ and observe the result.

\bigskip

1.  
If $Q_z$ is measurable and closed-valued, then by \jessie{Rockafellar VA 14.6}, we have a measurable selection $V \in \V_u$ such that $\inprod{z}{V(\omega)} = L'(u,\omega; z)$ for all $\omega \in \Omega$.
Thus, it just remains to show $Q_z$ is measurable and closed-valued.
First, we have $Q_z$ closed-valued by definition \jessie{Being lazy; will fill in later.}
%Now, we have $Q_z$ measurable if, for every open set $O \subset \reals^d$, the set $Q_z^{-1}(O) \in \B(\reals^d)$.
%\jessie{Rockafellar Wets VA Def 14.1}.
As measurable functions are closed under (countable) intersections, we have that $Q_z$ is the intersection of subgradients of $L$ and functions where the subgradient equals the directional derivative.
$V \in \V_u$ is measurable by definition.
Moreover, the function $\omega \mapsto \partial L(u,\omega)$ is measurable by \jessie{Rockafellar: Measurable Dependence, Corollary 4.6}, which also ensures $Q_z(\omega) \neq \emptyset$ for any $\omega \in \Omega$.
Now, we aim to show the set $\{c \in \reals^d: \inprod{z}{c} = L'(u,\omega;z)\}$ is measurable; as $L'$ is measurable and $\inprod{z}{\cdot}$ is continuous in its second argument, so the set is measurable. \jessie{Copying Russian argument}

As $Q_z$ is the intersection of these three sets, we conclude $Q_z$ is measurable and closed-valued, and therefore has a measurable selection $V \in \V_u$.

\bigskip

2.  
This statement follows trivially as $V(\omega) \in \partial L(u,\omega)$ for all $\omega \in \Omega$.

\iffalse
First, we want to show $\E_p V \in \E_p \partial L(u,Y)$.
That is, $\int V(\omega) dp\omega \in \int \partial L(u,\omega) dp\omega = \{\E_p V(Y) : V \textrm{ measurable, }\, V(\omega) \in \partial L(u,\omega)\}$.
Since $V$ is measurable, it is just left to show $V(\omega) \in \partial L(u,\omega)$ $p$-almost surely.

By construction, for all $\omega \in \Omega$, we have $\inprod{V(\omega)}{z} = L'(u,\omega;z)$.
This implies 
\begin{align*}
\inprod{V(\omega)}{z} &= L'(u,\omega;z)\\
\inprod{V(\omega)}{z} &= \sup_{x \in \partial L(u,\omega)} \inprod{x}{z} \jessie{\textrm{Boyd and Vandenberghe slides}}\\
\implies \inprod{V(\omega)}{z} &\geq \inprod{x}{z} \; \forall x \in \partial L(u,\omega)\\
z := \omega - t \text{\footnotemark}\implies \inprod{V(\omega)}{\omega - t} &\geq \inprod{x}{\omega - t} \geq L(u,\omega) - L(u,t) \; \forall x \in \partial L(u,\omega)\\
\implies \inprod{V(\omega)}{\omega-t} &\geq L(u,\omega) - L(u,t)\,\forall t \in \reals^d\\
V(\omega) &\in \partial L(u,\omega) \\ 
\end{align*}
\footnotetext{can we do this?  since $z$ is fixed but $t$ is a variable}
\jessie{This is actually a bit stronger than we need; is that part of my issue?}
\fi

\bigskip

3.
We want to show $\inprod{z}{v} = \E_p L'(u,Y,z)$.
This follows from linearity of expectation and construction of $V$; to see this, consider
\begin{align*}
\inprod{z}{v} &= \inprod{z}{\E_p V} \\
 &= \E_p \inprod{z}{V} \jessie{Re-evaluate}\\
 &= \int \inprod{z}{V(\omega)} dp\omega \\
 &= \int L'(u,\omega; z) dp\omega \\
 &= \E_p L'(u,Y;z)
\end{align*}
